DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-11-05 Paper_May_2012/main.tex
Paper: Section 2.1.2: Ein wenig geändert, allerdings fand ich die Vorarbeiten von Christian schon recht gut.

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[Discretization] Specify what kind of triangulation, finite element, quadrature rule, etc. we want to use.
\item[Problem-data] Define the weak form, the corresponding matrices, the functionals as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[Solvers] Which (non-)linear solver should we use? 
\mymarginpar{besser algorithmen?}
\item[Miscellaneous] ParameterReader? Outputhandler etc.
\end{description} 
[Hier eventuell auf Unterschied zw. blosser Objekterstellung und dem tatsaechlichen 'Programmieren' fuer localpde etc. eingehen.]

@@

\paragraph{\texttt{main.cc}}
In this file we gather the basic components that are necessary for the solution process of a (stationary) partial differential equation. The structure of the program is basically the same for every equation under consideration, what changes is typically just the \textit{kind} of  finite element or the linear solver.\marginpar{Bin mit diesen Saetzen nicht so ganz zufrieden, mir faellt aber gerade nichts besseres ein. Mir geht es hier darum, dass ich nicht wirklich was programmieren, sondern eher etwas auswaehle, wohingegen in den anderen files tatsaechlich was zu tun ist.}  
%The implementation of problem related data happens somwhere else, here we basically just pick together things like th

@@

Next, we create some quadrature rules (for the evaluation of integrals over cells and faces) and a finite element appropriate for the PDE at hand. We choose a gaussian quadrature of order three and the previosly mentioned Tayler-Hood element. \marginpar{Wie klar machen, dass vieles von deal.II uebernommen und nicht selbst geschrieben?}
\begin{lstlisting}
  QGauss<2> quad_formula(3);
  QGauss<1> face_quad_formula(3);
  FESystem<2> finite_element(dealii::FE_Q<2>(2), 2,
                             dealii::FE_Q(1), 1);
\end{lstlisting}
%IntegratorDataContainer<VECTOR, 2> idc(quad_formula,
%                                         face_quad_formula);
%The \texttt{IntegratorDataContainer} class just makes the constructor interface of 
Then, the \texttt{LocalPDE} and \texttt{LocalFunctional} objects come into play.
\begin{lstlisting}
LocalPDE<VECTOR> local_pde;
LocalFunctional<VECTOR> local_functional;
\end{lstlisting}
These hold the information of the PDE and functional under consideration. Further explication is given below because
the definition of this classes is done spererately in \texttt{localpde.h} resp.   \texttt{localfunctional.h}.

@@

In the next lines, we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{i,h}$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previosly selected quadrature formula. 
%     const FEValuesExtractors::Vector velocities(0);
%     const FEValuesExtractors::Scalar pressure(2); 
 \begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<n_dofs_per_cell;i++)
        local_cell_vector(i) += ( _nu 
                                * vgrads * phi_i_grads_v 
                                + //all the other terms)
                                * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.
% [CellEquation]
% 
% \begin{lstlisting}
%   void CellRightHandSide(const CellDataContainer<2>& cdc,
%                          dealii::Vector<double> &local_cell_vector,
%                          double scale);
% \end{lstlisting}
% 
% \begin{lstlisting}
%   void CellMatrix(const CellDataContainer<2>& cdc,
%                   dealii::FullMatrix<double> &local_entry_matrix);
% \end{lstlisting}
% The PDE under consideration is nonlinear, so let us briefly recapitulate manner such that for solving 
% the Laplace problem the following nonlinear step is necessary to recapitulate:
% Find $\delta u\in V$:
% \[
% (\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
% \]
% and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
% Therefore, the implementation reads for the right hand side:
% \begin{lstlisting}
%    void CellEquation(...)
%     {
%       // implement (\nabla u, \nabla \phi) 
%     }
% \end{lstlisting}
% Specifically, the inner loops reads
% \begin{lstlisting}
%    for (all quadrature points per cell q)
%       {
%         Tensor<2, 2> ugrads; // solution vector
%  
% 	for (all dofs per cell i)
%         {
% 	  const Tensor<2, 2> phi_i_grads_u; // test function
% 
%           local_cell_vector(i) += scale * 
%             scalar_product(ugrads, phi_i_grads_u) *
% 	    state_fe_values.JxW(q_point);
%         }
%       }
% \end{lstlisting}
% For the left hand side, we need to implement the Jacobian matrix:
% \begin{lstlisting}
%    void CellMatrix(...)
%     {
%       // implement  (\nabla \delta u, \nabla \phi)
%     }
% \end{lstlisting}
% Specifically,
% \begin{lstlisting}
%       for (all quadrature points per cell q)      
%         for (all dofs per cell i)    
%           for (all dofs per cell j)
%           {
%             local_entry_matrix(i, j) += 
%               scalar_product(phi_grads_u[j], phi_grads_u[i]) * 
%               state_fe_values.JxW(q_point);
%           }
% \end{lstlisting}
% The integration of the equation for each DoF and each quadrature point on each
% cell is then done 
% in the \texttt{templates/integrator.h} function. Finally, as usually in 
% finite element computations, this leads to a discrete system. Since
% we treat each problem with a Newton method, within each Newton step, the 
% following linear system must be solved:
% \[
% A\delta u = b
% \]
% in which $A$ is specified by \texttt{CellMatrix(...)} and 
% $b$ by \texttt{CellEquation(...)}. The vector $\delta u$ denotes the solution to the 
% present Newton step:
% \[
% u^n = u^{n-1} + \omega \delta u
% \]
% with a line search parameter $\omega\in (0,1]$.

>>>>

In the next section we will now sketch the way a user needs to implement the
user-dependent data as the local cell integrals and the initialization of the
objects described above.
\subsubsection{Problem dependent implementation of stationary PDE problems}
To solve a stationary PDE problem, the user has to provide some code which can roughly be divided into the following main sections:
\begin{description}
\item[Discretization] Specify what kind of triangulation, time stepping
  scheme, finite element, quadrature rules, etc. we want to use.
\item[Problem-data] Define the weak form, the corresponding matrices, the functionals as well as the boundary conditions. With exception of the boundary conditions, we have to specify all this only cell-wise, \dope{} handles the rest for us.
\item[Solvers] Which (non-)linear solver should we use? 
\item[Additional features] ParameterReader to read in the run-time parameters
  such as material coefficients or adjustments for the solvers like tolerance
  for Newton's method. Second, definition of Outputhandler etc.
\end{description}

@@

\paragraph{\texttt{main.cc}}
In this file we gather the basic components that are necessary for the
solution process of a (stationary) partial differential equation. The
structure of the program is basically the same for every equation under
consideration, what changes is typically just the \textit{kind} of  finite
element or the linear solver. Here, a key issue is that we only need to pick 
some pre-implemented routines from the \dope{} library rather than 
implementing new features.

@@

Next, we create some quadrature rules (for the evaluation of integrals over
cells and faces) and a finite element appropriate for the PDE at hand. We
choose a Gaussian quadrature of a certain degree and the previosly mentioned
Tayler-Hood element. These routines are quite standard in each finite element
software and we take advantage to take them from \deal{}.

%IntegratorDataContainer<VECTOR, 2> idc(quad_formula,
%                                         face_quad_formula);
%The \texttt{IntegratorDataContainer} class just makes the constructor interface of 
Then, the two characteristic features \texttt{LocalPDE} and \texttt{LocalFunctional} objects come into play.
\begin{lstlisting}
LocalPDE<VECTOR> local_pde;
LocalFunctional<VECTOR> local_functional;
\end{lstlisting}
These hold the information of the PDE and the functional under consideration. Further explication is given below because
the definition of this classes is done spererately in \texttt{localpde.h} resp.   \texttt{localfunctional.h}.

@@

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{i,h}$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previosly selected quadrature formula. 
%     const FEValuesExtractors::Vector velocities(0);
%     const FEValuesExtractors::Scalar pressure(2); 
 \begin{lstlisting}
    for(unsigned int q_point = 0; q_point < n_q_points; q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<n_dofs_per_cell;i++)
        local_cell_vector(i) += ( _nu 
                                * vgrads * phi_i_grads_v 
                                + //all the other terms)
                                * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.
% [CellEquation]
% 
% \begin{lstlisting}
%   void CellRightHandSide(const CellDataContainer<2>& cdc,
%                          dealii::Vector<double> &local_cell_vector,
%                          double scale);
% \end{lstlisting}
% 
% \begin{lstlisting}
%   void CellMatrix(const CellDataContainer<2>& cdc,
%                   dealii::FullMatrix<double> &local_entry_matrix);
% \end{lstlisting}
% The PDE under consideration is nonlinear, so let us briefly recapitulate manner such that for solving 
% the Laplace problem the following nonlinear step is necessary to recapitulate:
% Find $\delta u\in V$:
% \[
% (\nabla \delta u, \nabla \phi) = -(\nabla u, \nabla \phi) 
% \]
% and $u^{n+1} = u^n + \delta u$ for $n=1,2,3,ldots$.
% Therefore, the implementation reads for the right hand side:
% \begin{lstlisting}
%    void CellEquation(...)
%     {
%       // implement (\nabla u, \nabla \phi) 
%     }
% \end{lstlisting}
% Specifically, the inner loops reads
% \begin{lstlisting}
%    for (all quadrature points per cell q)
%       {
%         Tensor<2, 2> ugrads; // solution vector
%  
% 	for (all dofs per cell i)
%         {
% 	  const Tensor<2, 2> phi_i_grads_u; // test function
% 
%           local_cell_vector(i) += scale * 
%             scalar_product(ugrads, phi_i_grads_u) *
% 	    state_fe_values.JxW(q_point);
%         }
%       }
% \end{lstlisting}
% For the left hand side, we need to implement the Jacobian matrix:
% \begin{lstlisting}
%    void CellMatrix(...)
%     {
%       // implement  (\nabla \delta u, \nabla \phi)
%     }
% \end{lstlisting}
% Specifically,
% \begin{lstlisting}
%       for (all quadrature points per cell q)      
%         for (all dofs per cell i)    
%           for (all dofs per cell j)
%           {
%             local_entry_matrix(i, j) += 
%               scalar_product(phi_grads_u[j], phi_grads_u[i]) * 
%               state_fe_values.JxW(q_point);
%           }
% \end{lstlisting}
% The integration of the equation for each DoF and each quadrature point on each
% cell is then done 
% in the \texttt{templates/integrator.h} function. Finally, as usually in 
% finite element computations, this leads to a discrete system. Since
% we treat each problem with a Newton method, within each Newton step, the 
% following linear system must be solved:
% \[
% A\delta u = b
% \]
% in which $A$ is specified by \texttt{CellMatrix(...)} and 
% $b$ by \texttt{CellEquation(...)}. The vector $\delta u$ denotes the solution to the 
% present Newton step:
% \[
% u^n = u^{n-1} + \omega \delta u
% \]
% with a line search parameter $\omega\in (0,1]$.
